---
title: "Programming for Economists Group 6" 
author: "Daan Blok (2821386), Daniel ten Broeke (2823206), Dries A. M. Berns (2769645), Konstantinos Ntovas (2860092), Mariah P. Mendoza (2856166), Sofie A. Hopman (2745673), Vlad A. Partenie (2838992)" 
date: "`r Sys.Date()`" 
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) #makes the code not print in the pdf
```

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("scales")
install.packages("cbsodataR")
install.packages("rlang")
install.packages("sf")
install.packages("giscoR")
install.packages("tinytex")
#installing the packages
```

```{r packages, include=FALSE}
library(tidyverse)
library(scales)
library(cbsodataR)
library(rlang)
library(sf)
library(giscoR)
library(tinytex)
#utilising the packages
```

```{r defining constants and mapping}
province_map <- c(
  "pv20" = "Groningen", "pv21" = "Friesland", "pv22" = "Drenthe", "pv23" = "Overijssel",
  "pv24" = "Flevoland", "pv25" = "Gelderland", "pv26" = "Utrecht", "pv27" = "Noord-Holland",
  "pv28" = "Zuid-Holland", "pv29" = "Zeeland", "pv30" = "Noord-Brabant", "pv31" = "Limburg",
  "nl01" = "Nederland"
)

age_map <- c(
  "70100" = "0-19", "70200" = "0-19", "70300" = "0-19", "70400" = "0-19", "70500" = "20-39",
  "70600" = "20-39", "70700" = "20-39", "70800" = "20-39", "70900" = "40-59", "71000" = "40-59",
  "71100" = "40-59", "71200" = "40-59", "71300" = "60-79", "71400" = "60-79", "71500" = "60-79",
  "71600" = "60-79", "71700" = "80+", "71800" = "80+", "71900" = "80+", "22000" = "80+"
)

kenmerken_map <- c(
  "1014800" = "Owned Home", "1014950" = "Rented (No Benefit)", "1014900" = "Rented (With Benefit)"
)

years_to_plot <- c("2024jj00", "2014jj00", "2004jj00")

table_ids <- c("83625NED", "86004NED", "71488NED")

table_column_map <- list(
  "83625NED" = c("RegioS", "Perioden", "GemiddeldeVerkoopprijs_1"),
  "86004NED" = c("KenmerkenVanHuishoudens", "RegioS", "Perioden", "GemiddeldGestandaardiseerdInkomen_3", "Populatie"),
  "71488NED" = c("Geslacht", "Leeftijd", "RegioS", "Perioden", "TotaalPersonenInHuishoudens_1", "Alleenstaand_4", "TotaalSamenwonendePersonen_5")
)

plot_filters <- list(
  p1 = "RegioS == 'Nederland' & !is.na(Jaar) & !is.na(HuisPrijs)",
  p2 = "RegioS %in% province_map[1:12] & Perioden == '2023jj00' & !is.na(HuisPrijs) & !is.na(RegioS)",
  p3 = "RegioS == 'Nederland' & KenmerkenVanHuishoudens == 1050010 & !is.na(GestandaardiseerdInkomen) & !is.na(Jaar)",
  p4 = "RegioS %in% province_map[1:12] & Perioden == '2023jj00' & KenmerkenVanHuishoudens == 1050010 & !is.na(GestandaardiseerdInkomen) & !is.na(RegioS)",
  p5 = "RegioS == 'Nederland' & KenmerkenVanHuishoudens == 1050010 & !is.na(Jaar) & Jaar >= 2011 & !is.na(HousePriceIndex) & !is.na(IncomeLevelIndex)",
  p6 = "RegioS == 'Nederland' & KenmerkenVanHuishoudens == 1050010 & !is.na(Jaar) & Jaar >= 2011 & !is.na(Ratio)",
  p7 = "RegioS == 'Nederland' & is.na(KenmerkenVanHuishoudens) & is.na(Populatie) & Geslacht == 'T001038' & !is.na(OpZichzelfWonend) & !is.na(AgeGroup) & Jaar == 2024",
  p8 = "RegioS == 'Nederland' & !is.na(HousingType) & Geslacht == 'T001038' & !is.na(GestandaardiseerdInkomen) & !is.na(Jaar)",
  p9 = "RegioS == 'Nederland' & !is.na(Jaar) & !is.na(HuisPrijs)"
)

gis_to_regio <- c(
  "Groningen" = "Groningen", "Friesland (NL)" = "Friesland", "Drenthe" = "Drenthe", "Overijssel" = "Overijssel",
  "Flevoland" = "Flevoland", "Gelderland" = "Gelderland", "Utrecht" = "Utrecht", "Noord-Holland" = "Noord-Holland",
  "Zuid-Holland" = "Zuid-Holland", "Zeeland" = "Zeeland", "Noord-Brabant" = "Noord-Brabant","Limburg (NL)" = "Limburg"
)
#making vectors of the useful things such that the code is cleaner and easier to read
```

```{r defining custom themes}
my_theme <- theme_minimal() +
  theme(plot.title = element_text(hjust = 0), 
        plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
        axis.text.x = element_text(angle = 45, hjust = 1))

theme_map_clean <- theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        plot.margin = margin(t = 10, r = 10, b = 10, l = 10))
#defining themes for the plot for consistency
```

# Part 1 - Identifying a Social Problem

The housing market in the Netherlands is currently facing significant problems, especially for young people. There are long waiting lists for rental housing, and in addition, there are high rental prices in the private sector. The prices of owner-occupied homes have risen sharply in recent years, making it increasingly difficult for young people and young adults to find affordable housing (Wat Zijn Gevolgen van de Onzekere Woningmarkt Voor Jongeren? \| Nederlands Jeugdinstituut, n.d.-b). This has led to a historic low in the number of young people who have purchased a home. According to the CBS, there is a shortage of approximately 315,000 homes in the Netherlands (Wat Zijn Gevolgen van de Onzekere Woningmarkt Voor Jongeren? \| Nederlands Jeugdinstituut, n.d.-b). Furthermore the research by Wiesel et al. (2021) highlights that there are significant differences between generations regarding income inequality. Baby boomers have benefited from relatively affordable housing, while younger generations, such as Generation X and millennials, face greater challenges in covering housing costs. This has direct implications for their economic opportunities. The rising house prices and increasing housing costs have contributed to the growing inequality in the housing market. Households with middle incomes are also experiencing increasing problems. Many of these households fall between the cracks, as their income is too high for social rental housing but too low for the private rental market. This results in limited access to affordable housing (Schilder et al., n.d.). Although the social rental sector in the Netherlands offers affordable rents in absolute terms, the pressure on this sector is increasing, particularly due to the growing number of households that rely on social housing (Schilder et al., n.d.). The combination of these factors makes the situation in the housing market particularly urgent and calls for attention from policymakers to address the issues and ensure access to affordable housing for young people and other vulnerable groups.

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data, echo = TRUE}
load_cbs_raw <- function(table_id) {
  cbs_get_data(table_id, typed = FALSE)
}

raw_tables <- suppressWarnings(map(table_ids, load_cbs_raw))
#loads the datasets through their predefined ids
```

## 2.2 & 2.3 Short summary of the dataset and the variables included described

```{r printing the data, include=FALSE}
print(head(raw_tables)) #prints the raw datasets
```

Three external datasets are retrieved directly from CBS via the internet, using a predefined vector of dataset codes to automate access and ensure reproducibility. The specific codes used are listed in the corresponding vector `table_ids`. These datasets are compiled by CBS using a combination of integral administrative sources and, in some cases, survey data. For example, the Integrale inkomens- en vermogensstatistiek (Pouwels-Urlings, 2021) is based on comprehensive records from institutions such as the Belastingdienst (Dutch Tax Authority), Dienst Uitvoering Onderwijs (DUO), Uitvoeringsinstituut Werknemersverzekeringen (UWV), and the Basisregistratie Personen (BRP). The Huizenprijsindex bestaande koopwoningen (83625NED) relies on transaction data from the Kadaster (Centraal Bureau voor de Statistiek, 2008), which is updated monthly and considered immediately final. Meanwhile, the Stelsel van Sociaal-statistische Bestanden (Bakker et al., 2014) draws primarily on administrative registrations, but also incorporates data from the Enquête Beroepsbevolking (EBB – Labour Force Survey). These sources ensure the statistical accuracy and reliability of the datasets used in this analysis.

The first dataset contains three variables: `RegioS`, which denotes region codes defined in the dataset’s metadata and linked to the `province_map`; `Perioden`, which covers yearly data from 1995 to 2024, using the CBS format (e.g., 1995JJ00) that corresponds to calendar years; and `GemiddeldeVerkoopprijs_1`, which provides average sale prices of homes per region and year.

The second dataset is more detailed, with 20 variables related to household characteristics. It includes `RegioS` and `Perioden` beginning from 2011, and `KenmerkenVanHuishoudens`, a coded variable that identifies household types—filtered for relevant categories using `kenmerken_map`. The variable `Populatie` distinguishes subpopulations, such as private households with or without students. Additionally, it contains both absolute (`ParticuliereHuishoudens_1`) and relative (`ParticuliereHuishoudensRelatief_2`) household counts. Crucially, this dataset includes several income indicators, such as average and median standardised income (`GemiddeldGestandaardiseerdInkomen_3`, `MediaanGestandaardiseerdInkomen_4`) and income decile groups (`GestandaardiseerdInkomen1e10Groep_7` to `_16`), making it an essential resource for assessing income distribution in relation to housing affordability.

The third dataset complements the other two by focusing on the demographic composition of households and living arrangements. It includes 16 variables and covers the period from 2000 onward. The variable `Geslacht` distinguishes between all persons, men, and women using CBS codes (e.g., T001038, 3000, 4000). Age groups are captured via `Leeftijd`, defined in the metadata and referenced with the `age_map` vector. Like the others, it includes `RegioS` and `Perioden`. It offers a breakdown of household roles and living arrangements, including single-person households (`Alleenstaand_4`), cohabiting individuals, children living at home, and various couple structures (e.g., married/unmarried, with/without children). It also includes more specific roles such as single parents (`OuderInEenouderhuishouden_10`) and individuals in institutional households. This dataset enables detailed insight into who lives on themselves, how household compositions shift over time, and how these patterns differ by age and gender—important factors when evaluating housing needs and affordability.

Together, these three datasets are highly complementary. The first provides a direct measure of housing prices, the second supplies critical socioeconomic and income-related variables, and the third offers insight into demographic and household structure changes over time. Their shared use of `RegioS` and `Perioden` allows for integration across geographic and temporal dimensions, making them well-suited for a multidimensional analysis of housing affordability.

However, several limitations must be acknowledged. First, the time coverage is inconsistent: the first dataset spans 1995–2024, the second starts in 2011, and the third from 2000. This mismatch complicates long-term trend analysis, especially for integrated models. Second, not all time points have complete data—some values are missing or not filled in. Third, the datasets contain many variables that are irrelevant to the specific research focus and must be filtered out, increasing preprocessing time. Lastly, while administrative sources offer consistency, they may overlook informal living situations or alternative housing arrangements not captured in official registers. Despite these challenges, the richness, credibility, and compatibility of these datasets make them an excellent foundation for a thorough, data-driven analysis of housing affordability in the Netherlands.

# Part 3 - Quantifying

## 3.1 Data cleaning

```{r removing unnecessary variables/rows and standarising the values}
clean_cbs_data <- function(df, table_id) {
  df <- df %>%
    mutate(across(c(RegioS, Perioden), ~ str_trim(tolower(as.character(.)))))
  
  colnames(df) <- str_trim(colnames(df))
  
  keep_cols <- table_column_map[[table_id]]
  
  keep_cols <- union(keep_cols, c("RegioS", "Perioden"))
  
  df <- df %>% select(any_of(keep_cols)) %>%
    filter(RegioS %in% tolower(names(province_map))) %>%
    {
      if ("Geslacht" %in% names(.)) filter(., Geslacht == "T001038") 
      else if ("Populatie" %in% names(.)) filter(., Populatie == "1050010") else .
    }
  
  return(df)
}

cleaned_tables <- map2(raw_tables, table_ids, clean_cbs_data) 
#cleaning the data a bit before merging for efficieny

```

```{r merging the data}
merged_df <- cleaned_tables %>% #merging the datasets
  reduce(full_join, by = c("RegioS", "Perioden")) %>%
  mutate(across(where(is.character), str_trim))
```

```{r detecting numeric columns and converting them to numbers}
numeric_like_cols <- names(merged_df)[
  sapply(merged_df, function(col) mean(str_detect(col, "^\\s*[0-9,.]+\\s*$"), na.rm = TRUE) > 0.8)
]

merged_df <- merged_df %>%
  mutate(across(all_of(numeric_like_cols), ~ parse_number(., na = c("", ".", "n.v.t.", "onbekend")))) 
#to check if the columns are numeric
```

```{r Transforming columns }
merged_df <- merged_df %>% #columns for easier computations
  mutate(
    RegioS = recode(RegioS, !!!province_map),
    Jaar = as.integer(str_sub(Perioden, 1, 4)),
    HuisPrijs = GemiddeldeVerkoopprijs_1 / 1000,
    GestandaardiseerdInkomen = GemiddeldGestandaardiseerdInkomen_3,
    AgeGroup = age_map[as.character(Leeftijd)],
    HousingType = kenmerken_map[as.character(KenmerkenVanHuishoudens)],
    YearStr = substr(Perioden, 1, 4),
  )
```

To integrate multiple CBS datasets for analysis, we applied a structured cleaning and merging process using `dplyr` from the Tidyverse. Each dataset was first standardised by trimming whitespace, converting text to lowercase, and filtering for consistent identifiers (`RegioS`, `Perioden`) and relevant subgroups (e.g., total population or total by gender). This ensured alignment across tables and enabled accurate joins.

The `full_join` merge was chosen to preserve all observations across datasets, minimising data loss due to missing values in specific tables. This was the optimal choice given the diverse content of each dataset and the need for a comprehensive merged view. In an ideal setting where all three datasets started in the same year—or at least before 2007 to capture the 2008 financial crisis for event analysis—merging would have been more straightforward, potentially allowing for more efficient `inner_join` and better integrity validation. The `full_join` was particularly necessary here due to the differences in the starting years of the datasets. A harmonised national data repository with consistent keys and metadata across tables would further improve merge quality and analytical coherence.

Each dataset was then subset to retain only relevant columns using a mapping list (`table_column_map`), making the merged output cleaner and more efficient for further analysis. Additionally, subpopulation filters (e.g., total population or total gender) were applied conditionally based on variable presence. This logic ensured consistency across years and tables, while minimising manual intervention.

Several types of data errors were identified and addressed systematically. First, numerous numeric variables were stored as strings and sometimes included invalid characters (e.g., ".", "n.v.t.", "onbekend"). To detect these issues, we scanned columns for numeric-like patterns and applied `parse_number()` with custom `na` values, which successfully recovered quantitative fields for analysis. Furthermore, Constants such as `province_map`, `age_map`, and `kenmerken_map` were defined at the top of the script to convert cryptic CBS codes into interpretable labels. This allowed for the generation of meaningful variables like `AgeGroup`, `HousingType`, and readable province names. Derived variables like `Jaar` (from `Perioden`) and `HuisPrijs` (scaled to thousands) were added to facilitate time-series plotting and unit consistency.

Finally, a list of filter expressions (`plot_filters`) was applied programmatically to subset the data for specific plots and analyses. These filters, defined as character strings and parsed using `parse_expr()`, allowed us to modularly apply consistent inclusion criteria across visualisations, further enforcing clarity and reproducibility.

## 3.2 Generating necessary variables

```{r generating necessary variables, include=FALSE}
reference_price_2011 <- merged_df %>% #filters data for the price in 2011
  filter(RegioS == "Nederland", Jaar == 2011) %>%
  pull(HuisPrijs) %>%
  first()

reference_income_2011 <- merged_df %>% #filters data for the income in 2011
  filter(RegioS == "Nederland", Jaar == 2011) %>%
  pull(GestandaardiseerdInkomen) %>%
  first()

merged_df <- merged_df %>% #creates four variables for further plots
  mutate(    OpZichzelfWonend = 100 * (Alleenstaand_4 + TotaalSamenwonendePersonen_5) 
             /      TotaalPersonenInHuishoudens_1,
             HousePriceIndex = 100 * HuisPrijs / reference_price_2011,
             IncomeLevelIndex = 100 * GestandaardiseerdInkomen / reference_income_2011,
             Ratio = HousePriceIndex / IncomeLevelIndex)

```

To analyse trends in housing affordability in the Netherlands, we constructed several new variables using the `mutate()` function in R, which allows for efficient transformation and combination of existing variables.

First, reference values for house prices and standardised income in 2011 were created by filtering the dataset for the national level `(RegioS == "Nederland")` and base year `(Jaar == 2011)`. These values serve as a benchmark for the index construction that follows.

The `HousePriceIndex` was calculated by dividing the nominal house price `(HuisPrijs)` by the national average in `2011` and multiplying by `100`. Similarly, the `IncomeLevelIndex` was created by dividing the standardised income `(GestandaardiseerdInkomen)` by its `2011` national reference and scaling by `100`. These indices allow for relative comparisons of housing prices and income levels over time, making temporal trends more interpretable regardless of absolute price or income levels.

The `Ratio` variable was constructed by dividing the `HousePriceIndex` by the `IncomeLevelIndex` and multiplying by `100`. This `Ratio` captures changes in housing affordability: values greater than 100 indicate that housing prices have increased more rapidly than income since 2011, suggesting a decline in affordability. This variable is central to our analysis, as it directly quantifies the relationship between income and housing costs.

Finally, the variable `OpZichzelfWonend` represents the share of individuals living independently. It was computed by summing individuals who live alone `(Alleenstaand_4)` and those in cohabiting households `(TotaalSamenwonendePersonen_5)`, dividing this by the total household population `(TotaalPersonenInHuishoudens_1)`, and multiplying by `100`. This measure is informative for examining demographic shifts in household composition, which may influence housing demand and affordability.

These constructed variables will be used to visualise trends over time, and evaluate how housing costs have evolved relative to income—providing critical insights into the changing landscape of housing affordability in the Netherlands.

```{r applying the list with filters on the plots}
merged_df <- map(plot_filters, ~ merged_df %>% 
                  filter(!!parse_expr(.x))) #applies the needed filters on the plots from a predetermined vector
```

## 3.3 Visualise temporal variation

```{r visualise temporal variation, out.width="52%", out.height="30%"}
p1 <- merged_df$p1 %>% #plots two temporal variation graphs
  ggplot(aes(x = Jaar, y = HuisPrijs)) +
  geom_line(color = "blue", linewidth = 1.2) +
  scale_x_continuous(breaks = seq(1995, 2024, by = 2)) +
  labs(title = "Average House Price (Netherlands)", x = "Year", y = "Price (EUR x 1,000)")

list(p1) %>% walk(~ print(.x + my_theme))

p3 <- merged_df$p3 %>%
  ggplot(aes(x = Jaar, y = GestandaardiseerdInkomen)) +
  geom_line(color = "blue", linewidth = 1.2) +
  scale_x_continuous(breaks = seq(2011, 2023, by = 2)) +
  labs(title = "Average Standardised Income (Netherlands)", x = "Year", y = "Income (EUR x 1,000)")

list(p3) %>% walk(~ print(.x + my_theme))
```

The affordability of housing in the Netherlands has undergone significant changes over time, as illustrated by the five temporal visualisations. Figure 1 presents the development of average house prices from 1995 to 2023, showing a steady upward trend, with particularly sharp increases after 2015. Figure 2 complements this by depicting the evolution of average standardised disposable income since 2011, which shows only moderate growth in comparison.

```{r visualise temporal variation two, out.width="52%", out.height="30%"}
p5 <- merged_df$p5 %>% #plots two temporal variation graphs
  ggplot(aes(x = Jaar)) +
  geom_line(aes(y = HousePriceIndex, color = "House Prices"), linewidth = 1.2) +
  geom_line(aes(y = IncomeLevelIndex, color = "Standardised Income"), linewidth = 1.2) +
  scale_x_continuous(breaks = seq(2011, 2023, by = 2)) +
  scale_color_manual(values = c("House Prices" = "darkred", "Standardised Income" = "darkgreen")) +
  labs(title = "House Price vs Standardised Income (Netherlands)", x = "Year", y = "Index (2011 = 100)", color = "Measure")

p6 <- merged_df$p6 %>%
  ggplot(aes(x = Jaar, y = Ratio)) +
  geom_line(color = "purple", linewidth = 1.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray") +
  scale_x_continuous(breaks = seq(2011, 2023, by = 2)) +
  labs(title = "Ratio of House Prices to Standardised Income", x = "Year", y = "Ratio")

list(p5, p6) %>% walk(~ print(.x + my_theme))
```

To better understand relative changes, Figure 3 presents both house prices and income as indexed values (2011 = 100). This comparison highlights that house prices have increased at a substantially faster rate than incomes, especially after 2013, suggesting a growing mismatch between the two variables. Finally, Figure 4 visualises the ratio of house prices to standardised income over time. The upward trajectory of this ratio indicates a declining affordability of housing, as house prices increasingly outpace income levels. The dashed reference line at a ratio of 1 serves as a benchmark: values above this threshold imply deteriorating affordability.

```{r visualise temporal variation three, out.width="52%", out.height="30%", fig.align='center'}
p8 <- merged_df$p8 %>% #plots a temporal variation graph
  group_by(Jaar, HousingType) %>%
  summarise(AverageIncome = mean(GestandaardiseerdInkomen, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Jaar, y = AverageIncome, color = HousingType, linetype = HousingType)) +
  geom_line(linewidth = 1.2) +
  scale_x_continuous(breaks = seq(2011, 2023, by = 2)) +
  labs(title = "Average Standardised Income by Housing Type", x = "Year", y = "Income (EUR x 1,000)", color = "Housing Type", linetype = "Housing Type")

list (p8) %>% walk(~ print(.x + my_theme))
```

The fifth visualisation presents average standardised income by housing tenure over the period 2011–2023. While younger individuals typically start their careers with lower incomes, this plot reveals that although the relative income gap between housing types has remained stable, the absolute difference in income levels between homeowners and renters has grown substantially over the past twelve years. This growing absolute gap exacerbates affordability challenges for lower-income renters, many of whom are young adults or households yet to enter the property market.

These visualisations align closely with the research topic on housing affordability, providing clear evidence of a growing temporal gap between income and house prices. Time is consistently displayed on the x-axis, ensuring the temporal dimension is appropriately represented.

## 3.4 Visualise spatial variation

```{r visualise spatial variation, out.width="52%", out.height="30%"}
#pulls the map of the Netherlands for spatial visualisation
nl_map <- gisco_get_nuts(year = "2021",nuts_level = 2,country = "NL",resolution = "01",spatialtype = "RG") %>%
  st_transform(crs = 28992) %>%
  mutate(RegioS = recode(NAME_LATN, !!!gis_to_regio)) %>%
  filter(!is.na(RegioS))

nl_map_p2 <- left_join(nl_map, distinct(merged_df$p2, RegioS, .keep_all = TRUE), by = "RegioS")
nl_map_p4 <- left_join(nl_map, distinct(merged_df$p4, RegioS, .keep_all = TRUE), by = "RegioS")

#plots the spatial variation graphs
p2 <- ggplot(nl_map_p2) +
  geom_sf(aes(fill = HuisPrijs), color = "white", size = 0.2) +
  scale_fill_gradient(low = "skyblue", high = "darkblue", name = "Price (EUR x 1,000)") +
  labs(title = "Average House Prices by Province (2023)") 

p4 <- ggplot(nl_map_p4) +
  geom_sf(aes(fill = GestandaardiseerdInkomen), color = "white", size = 0.2) +
  scale_fill_gradient(low = "skyblue", high = "darkblue", name = "Income (EUR x 1,000)") +
  labs(title = "Average Standardised Income by Province (2023)")

list(p2, p4) %>% walk(~ print(.x + theme_map_clean))
```

The spatial visualisations above demonstrate how the problem of housing affordability varies across the provinces of the Netherlands in 2023. Figure 1 shows average house prices by province, indicating that housing is most expensive in the Randstad area—particularly in North Holland, South Holland, and Utrecht—while prices are relatively lower in peripheral provinces such as Groningen, Friesland, and Drenthe.

Figure 2 displays the average standardised disposable income across the same provinces. Although higher income levels are also concentrated in the western part of the country, the regional relative variation in income is notably smaller than the relative variation in house prices. This imbalance implies that housing affordability issues are most severe in provinces where house prices significantly exceed income levels.

This visualisation aligns with the topic of housing affordability by highlighting the spatial mismatch between house prices and incomes, helping to identify which regions face the greatest challenges. Mapping these indicators at the provincial level provides a clear geographic perspective on affordability, supporting evidence-based analysis of regional disparities.

## 3.5 Visualise sub-population variation

```{r visualise subpopulations, out.width="52%", out.height="30%",fig.align='center'}
p7 <- merged_df$p7 %>% #plots the subpopulation boxplot
  mutate(Year = factor(YearStr, levels = sort(unique(substr(years_to_plot, 1, 4))))) %>%
  ggplot(aes(x = AgeGroup, y = OpZichzelfWonend, fill = AgeGroup)) +
  geom_boxplot() +
  labs(title = "Share of Individuals Living Indepent by Age Group", x = "Age Group (Years)", y = "Individuals Living Indepent (%)")


list(p7) %>% walk(~ print(.x + my_theme))
```

The boxplot visualises the distribution of the percentage of individuals living independently across five broad age groups. The 40–59 and 60–79 age groups show the highest and most consistent rates of independent living, with medians close to or above 90%. In contrast, the 0–19 age group has an extremely low share, reflecting that nearly all individuals in this group live dependently. The 20–39 and 80+ groups exhibit wider variability: for 20–39-year-olds, this suggests a transition period where many begin living independently at differing rates; for the 80+ group, the spread may reflect differences in health, support needs, or institutional living arrangements. This visualisation aligns closely with the topic of housing affordability by demonstrating that the problem manifests unevenly across sub-populations defined by age and housing tenure. The grouped plot and the selected variables effectively highlight how demographic and economic factors contribute to differential affordability pressures within the Dutch housing market.

## 3.6 Event analysis

```{r event-analysis, out.width="52%", out.height="30%", fig.align='center'}
p9 <- merged_df$p9 %>% #plots the event analysis graph
  ggplot(aes(x = Jaar, y = HuisPrijs, color = RegioS)) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_vline(xintercept = 2008, linetype = "dotted", color = "black", linewidth = 1) +
  scale_x_continuous(breaks = seq(1995, 2024, by = 2)) +
  labs(title = "Average House Price (Netherlands)", x = "Year", y = "Price (EUR x 1,000)", color = "Region")

list(p9) %>% walk(~ print(.x + my_theme))
```

The event analysis visualisation depicts average house prices in the Netherlands over time, with a specific focus on the year 2008, marked by a vertical dotted line. This year corresponds to the global financial crisis, a significant event that had a marked impact on housing markets worldwide.

Before 2008, house prices exhibited a steady upward trend. However, the visualisation shows a clear disruption around this event: house prices plateaued and even declined briefly following 2008, reflecting the crisis’s immediate effect on the housing market. After this period, prices resumed their upward trajectory, but with increased volatility and sharper increases in subsequent years.

This visualisation is directly relevant to the topic of housing affordability, as it highlights how a major economic event can temporarily alter housing price dynamics. The post-crisis acceleration in prices, despite income growth remaining comparatively modest, suggests that affordability pressures intensified after 2008. Understanding such event-driven changes is critical for analysing the temporal fluctuations and risks inherent in the housing market.

# Part 4 - Discussion

The results presented in the previous sections offer a comprehensive and multidimensional view of housing affordability in the Netherlands. Through the combination of temporal, spatial, and demographic analyses, several critical patterns emerge that warrant closer reflection.

First, the temporal visualisations clearly show a persistent and widening gap between house prices and standardised income over the last decade. While house prices have increased substantially since 2013, average standardised income has only grown modestly. This decoupling of income from housing costs has resulted in a steadily rising affordability ratio, particularly after 2015, indicating that housing is becoming less accessible for many Dutch residents. These findings support the concerns highlighted in Part 1, where growing generational inequality and the increasing economic marginalisation of middle-income households were described. The index-based approach `(2011 = 100)` proves useful in quantifying the speed and direction of these divergent trends over time.

The spatial dimension further emphasises the imbalance. House prices in the Randstad provinces (North Holland, South Holland, Utrecht) are significantly higher than in peripheral regions, yet the variation in income across provinces is much narrower. This implies that affordability problems are particularly acute in urban and economically dynamic regions, where housing demand outpaces supply. These spatial discrepancies are consistent with the national housing shortage identified by CBS and raise concerns about regional inequality and spatial segregation.

Demographically, the sub-population analysis reveals that affordability constraints affect age groups differently. Younger age cohorts (20–39) show considerable variation in independent living, suggesting that structural barriers—such as unaffordable housing—may be delaying household formation and economic independence. This demographic stagnation, especially among young adults, aligns with existing literature on declining homeownership rates and delayed life course transitions. The high consistency of independent living in middle-aged groups underscores that affordability challenges are disproportionately burdening newer market entrants.

The event analysis adds further nuance by highlighting the impact of the 2008 global financial crisis. While house prices temporarily stabilised or declined during this period, the recovery phase was marked by accelerated price increases, without a commensurate rise in income. This pattern suggests a long-term structural shift rather than a cyclical fluctuation. Moreover, the visual break in the trend underscores the vulnerability of the housing market to external shocks, reinforcing the need for resilience-oriented housing policy.

While the findings are robust, several limitations should be acknowledged. The inconsistent temporal coverage across datasets reduces the potential for long-term integrated analysis, particularly prior to 2011. Furthermore, administrative data, while rich and reliable, may fail to capture informal housing arrangements or underrepresented subpopulations such as undocumented migrants or those in non-traditional living setups. Additionally, the income measures used do not account for wealth accumulation (e.g., equity from homeownership), which may further differentiate affordability across households.

Nevertheless, the constructed indices and affordability ratio offer powerful tools for quantifying and comparing affordability over time, across regions, and between demographic groups. They provide a rigorous empirical foundation for policy evaluation and highlight where interventions may be most needed—namely in urbanised provinces, among young adults, and in the private rental sector where income constraints are most acute.

Future research could incorporate micro-level data on housing expenditures, debt burdens, or wealth to enrich the analysis. Moreover, evaluating the role of government policies (e.g., housing subsidies, rent control, zoning regulations) could offer valuable insights into potential levers for improving affordability.

In summary, the Dutch housing market has become increasingly unaffordable over the past decade, particularly for younger and lower-income populations. The widening gap between housing costs and incomes, compounded by spatial and demographic disparities, presents a pressing social and economic challenge. Addressing this issue will require a combination of data-informed policy reforms and targeted investments to restore affordability and equity in the housing system.

# Part 5 - Reproducibility

## 5.1 Github repository link

Github link: <https://github.com/Daan-Blok/ProgrammingForEconomists.git>

## 5.2 Reference list

Wat zijn gevolgen van de onzekere woningmarkt voor jongeren? \| Nederlands Jeugdinstituut. (n.d.-b). <https://www.nji.nl/op-jezelf-wonen/wat-zijn-gevolgen-van-de-onzekere-woningmarkt-voor-jongeren>

Wiesel, I., Ralston, L., & Stone, W. (2021). Understanding after-housing disposable income effects on rising inequality. Housing Studies, 38(2), 290–306. <https://doi.org/10.1080/02673037.2021.1882661>

Schilder, F., Scherpenisse, R., PBL Netherlands, & Tiwos. (n.d.). Affordable housing in the Netherlands. <https://www.pbl.nl/sites/default/files/downloads/PBL2018_Policy-and-practice-affordable-housing-in-the-Netherlands_3336_0.pdf>

Pouwels-Urlings, N. (2021, April 20). *Herziening van de Vermogensstatistiek 2006*. Centraal Bureau Voor De Statistiek. <https://www.cbs.nl/nl-nl/longread/discussion-papers/2021/herziening-van-de-vermogensstatistiek-2006> (86004NED)

*CBS dataportaal*. (n.d.-c). <https://opendata.cbs.nl/statline/portal.html?_la=nl&_catalog=CBS&tableId=86004NED&_theme=310#>

Centraal Bureau voor de Statistiek. (2008). Prijsindex bestaande koopwoningen methodebeschrijving. In *Prijsindex Bestaande Koopwoningen Methodebeschrijving*. <https://www.cbs.nl/-/media/_pdf/2018/08/methodebeschrijving-prijsindex-bestaande-koopwoningen.pdf> (83625NED)

*CBS dataportaal*. (n.d.). <https://opendata.cbs.nl/statline/portal.html?_la=nl&_catalog=CBS&tableId=83625NED&_theme=398#>

Bakker, B. F., Van Rooijen, J., & Van Toor, L. (2014). The System of social statistical datasets of Statistics Netherlands: An integral approach to the production of register-based social statistics. *Statistical Journal of the IAOS*, *30*(4), 411–424. <https://doi.org/10.3233/sji-140803> (71488ned)

*CBS dataportaal*. (n.d.-b). <https://opendata.cbs.nl/statline/portal.html?_la=nl&_catalog=CBS&tableId=71488ned&_theme=281#>
