---
title: "PFE group 6"
output: html_document
date: "2025-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Part 1 - Identify a Social Problem**

The housing market in the Netherlands is currently facing significant problems, especially for young people. There are long waiting lists for rental housing, and in addition, there are high rental prices in the private sector. The prices of owner-occupied homes have risen sharply in recent years, making it increasingly difficult for young people and young adults to find affordable housing (Wat Zijn Gevolgen van de Onzekere Woningmarkt Voor Jongeren? \| Nederlands Jeugdinstituut, z.d.-b). This has led to a historic low in the number of young people who have purchased a home. According to the CBS, there is a shortage of approximately 315,000 homes in the Netherlands (CBS Statline, z.d.). The research by Wiesel et al. (2021) highlights that there are significant differences between generations regarding income inequality. Baby boomers have benefited from relatively affordable housing, while younger generations, such as Generation X and millennials, face greater challenges in covering housing costs. This has direct implications for their economic opportunities. The rising house prices and increasing housing costs have contributed to the growing inequality in the housing market. Households with middle incomes are also experiencing increasing problems. Many of these households fall between the cracks, as their income is too high for social rental housing but too low for the private rental market. This results in limited access to affordable housing (Schilder et al., z.d.-b). Although the social rental sector in the Netherlands offers affordable rents in absolute terms, the pressure on this sector is increasing, particularly due to the growing number of households that rely on social housing (Schilder et al., z.d.-b). The combination of these factors makes the situation in the housing market particularly urgent and calls for attention from policymakers to address the issues and ensure access to affordable housing for young people and other vulnerable groups.

**This leads to the following social problem: Access to affordable housing for vulnerable groups.**

## Part 2.3 – Describe the Type of Variables

The data sets used in this project contain variables related to socioeconomic status (SES), such as income levels and housing prices, across different municipalities in the Netherlands. These are some examples:

- **Median Income (from data set 71488NED)** – A numeric variable representing the median standardized income per household.  
- **Average Housing Price (from data set 83625NED)** – A numeric variable showing the average transaction price of homes.  
- **Housing Stock by Type (from data set 86004NED)** – Numeric variables showing the number of dwellings, categorized by ownership and dwelling type.  
- **Municipality** – A categorical variable used to identify the region for all observations.  
- **Year** – A time-based variable indicating when the data was collected.  

### Type of Information

The variables in these data sets provide **socioeconomic status (SES)** information. They do **not** contain health data.

### How the Data Was Collected

The data sets are compiled from **administrative sources**, such as:

- The Dutch Tax Authority (for income data)  
- The Land Registry (for housing prices)  
- Local municipal records (for housing stock)  

This means that the data was **not collected through interviews or surveys**, but from official government databases and registries. This method ensures consistency and reliability across all municipalities.

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
